### 2018-03-07 chapter2/ms.tex

\section*{Abstract}
The first direct detection of gravitational waves (GW) by the ground-based Advanced LIGO/Virgo interferometers is expected to occur within the next few years.  These interferometers are designed to detect the mergers of compact object binaries composed of neutron stars and/or black holes to a fiducial distance of $\sim 200$ Mpc and a localization region of $\sim 100$ deg$^2$.  To maximize the science gains from such GW detections it is essential to identify electromagnetic (EM) counterparts.  Among the wide range of proposed counterparts, the most promising is optical/IR emission powered by the radioactive decay of $r$-process elements synthesized in the neutron-rich merger ejecta -- a ``kilonova''.  Here we present detailed simulated observations that encompass a range of strategies for kilonova searches during GW follow-up.  We utilize these simulations to assess both the detectability of kilonovae and our ability to distinguish them from a wide range of contaminating transients in the large GW localization regions.  We find that if pre-existing deep template images for the GW localization region are available, then nightly observations to a depth of $i\approx 24$ mag and $z\approx 23$ mag are required to achieve a 95\% detection rate; observations that commence within $\sim 12$ hours of trigger will also capture the kilonova peak and provide stronger constraints on the ejecta properties.  We also find that kilonovae can be robustly separated from other known and hypothetical types of transients utilizing cuts on color ($i-z\gtrsim 0.3$ mag) and rise time ($t_{{\rm rise}}\lesssim 4$ days).  In the absence of a pre-existing template the observations must reach $\sim 1$ mag deeper to achieve the same kilonova detection rate, but robust rejection of contaminants can still be achieved.  Motivated by the results of our simulations we discuss the expected performance of current and future wide-field telescopes in achieving these observational goals, and find that prior to LSST the Dark Energy Camera (DECam) on the Blanco 4-m telescope and Hyper Suprime-Cam (HSC) on the Subaru 8-m telescope offer the best kilonova discovery potential.

@@

Extensive theoretical and observational work has been performed to understand the potential EM emission from a merging compact object binary. The range of potential counterparts covers the entire EM spectrum. The most popular counterpart is a short-hard gamma-ray burst \citep[SGRB; e.g.,][]{Paczynski1986,Narayan+92,Berger2014} powered by rapid accretion onto the compact remnant which leads to a strongly beamed relativistic jet. The evidence for a connection between SGRBs and compact object mergers includes the association with older stellar populations than for long GRBs \citep{Berger+05,Bloom+06,Berger2011,Fong+11,Fong+13}, large physical offsets between the burst locations and their host galaxies indicative of progenitor kicks \citep{Fong+10, FongBerger13}, weak association with the underlying stellar light of their hosts \citep{Fong+10, FongBerger13}, and a lack of association with supernovae \citep{Fox+05,Soderberg+06,Berger2010,FongBerger13,Berger2014}. The relativistic jet can also be observed in the x-ray, optical, and radio as non-thermal afterglow emission arises from interaction with the ambient medium. This emission is also affected by relativistic beaming but may still be seen off-axis as the jet decelerates and expands. In particular, for large off-axis angles $(\theta\gtrsim 2 \theta_j)$ the emission will be most pronounced in the radio on timescales of weeks to years \citep{NakarPiran11,MetzgerBerger12}.

@@

To help provide a framework for these observations, and to more broadly investigate the coupled issues of kilonova detectability and contaminant rejection, we present here a systematic study of the feasibility of kilonova detections and assessing the effects of a population of contaminant sources during follow-up of a GW trigger using wide-field optical telescopes. This study will guide the development of observational strategies for wide-field searches that maximize the likelihood of detecting a GW optical counterpart. The paper is organized as follows: In \cref{sec:ch2_contaminants} we describe the kilonova models utilized in this work and develop the list of contaminating transients considered in this work. In \cref{sec:ch2_phase} we  describe observational features that may be used to isolate kilonovae from the population of potential contaminants. In \cref{sec:ch2_MCsims} we present detailed Monte Carlo simulations of observations to investigate the detectability of kilonovae and contaminant rejection for various observational strategies. In \cref{sec:ch2_diff}, we address issues that arise from the possible lack of a deep pre-existing template. We then discuss the effects of modifications to the standard kilonovae models on our results in \cref{sec:ch2_altkilo}. In \cref{sec:ch2_disc}, we asses the expected performance of current and future wide-field telescopes in the framework of kilonova detectability and contaminant rejection and develop strategies for GW follow-up. Conclusions are presented in \cref{sec:ch2_conc}.

@@

\section{Kilonovae and Contaminants}
\label{sec:ch2_contaminants}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{./figs/chapter2/ch2_f1.pdf}
\caption{Top: Kilonova spectra from \citet{BarnesKasen13} at $\sim$ 1 d and $\sim$ 1 week with ejecta properties of $\Mej =10^{-2}\;M_{\odot}$ and $\beta_{\text{ej}} = 0.2$. Bottom: Light curve for the same model parameters computed at a fiducial distance of 200 Mpc in the DECam {\em r}, {\em i}- and {\em z}-band filters. \vspace{10pt}}
\label{fig:ch2_kilosample}
\end{figure}

@@

@@

@@

\begin{figure}[h!]
\centering
\includegraphics[trim= 0 0 0 240 ,clip,width=0.6\textwidth]{./figs/chapter2/ch2_f2.pdf}
\caption{Top: {\em i}-band light curves of the contaminants considered in this work at a fiducial distance of 200 Mpc. Bottom: {\em z}-band light curves. See \cref{sec:ch2_WDmerge,sec:ch2_AIC,sec:ch2_ELDD,sec:ch2_type.Ia,sec:ch2_PSFast,sec:ch2_typeIa} for details.}
\label{fig:ch2_LCcont}
\end{figure}

>>>>

\clearpage
\section*{Abstract}
The first direct detection of gravitational waves (GW) by the ground-based Advanced LIGO/Virgo interferometers is expected to occur within the next few years. These interferometers are designed to detect the mergers of compact object binaries composed of neutron stars and/or black holes to a distance of $\sim 200$ Mpc and a localization region of $\sim 100$ deg$^2$.  To maximize the science gains from GW detections it is essential to identify electromagnetic (EM) counterparts. The most promising proposed counterpart is optical/IR emission powered by the radioactive decay of $r$-process elements synthesized in the merger ejecta -- a ``kilonova''.  Here we present simulated observations that encompass a range of strategies for kilonova searches during GW follow-up.  We assess both the detectability of kilonovae and our ability to distinguish them from a wide range of contaminating transients. We find that if pre-existing deep template images are available, then nightly observations to a depth of $i\approx 24$ mag and $z\approx 23$ mag are required to achieve a 95\% detection rate; observations that commence within $\sim 12$ hours of trigger will also capture the kilonova peak and provide stronger constraints on the ejecta properties.  We also find that kilonovae can be robustly separated from other contaminating transients with cuts on color ($i-z\gtrsim 0.3$ mag) and rise time ($t_{{\rm rise}}\lesssim 4$ days).  In the absence of pre-existing templates the observations must reach $\sim 1$ mag deeper to achieve the same kilonova detection rate, but robust rejection of contaminants is still possible.  Motivated by these results we discuss the expected performance of current and future wide-field telescopes in achieving these observational goals finding that prior to LSST the Dark Energy Camera (DECam) and Hyper Suprime-Cam (HSC) offer the best kilonova discovery potential.

@@

\clearpage
Extensive theoretical and observational work has been performed to understand the potential EM emission from a merging compact object binary. The range of potential counterparts covers the entire EM spectrum. The most popular counterpart is a short-hard gamma-ray burst \citep[SGRB; e.g.,][]{Paczynski1986,Narayan+92,Berger2014} powered by rapid accretion onto the compact remnant which leads to a strongly beamed relativistic jet. The evidence for a connection between SGRBs and compact object mergers includes the association with older stellar populations than for long GRBs \citep{Berger+05,Bloom+06,Berger2011,Fong+11,Fong+13}, large physical offsets between the burst locations and their host galaxies indicative of progenitor kicks \citep{Fong+10, FongBerger13}, weak association with the underlying stellar light of their hosts \citep{Fong+10, FongBerger13}, and a lack of association with supernovae \citep{Fox+05,Soderberg+06,Berger2010,FongBerger13,Berger2014}. The relativistic jet can also be observed in the x-ray, optical, and radio as non-thermal afterglow emission arises from interaction with the ambient medium. This emission is also affected by relativistic beaming but may still be seen off-axis as the jet decelerates and expands. In particular, for large off-axis angles $(\theta\gtrsim 2 \theta_j)$ the emission will be most pronounced in the radio on timescales of weeks to years \citep{NakarPiran11,MetzgerBerger12}.

@@

\clearpage
To help provide a framework for these observations, and to more broadly investigate the coupled issues of kilonova detectability and contaminant rejection, we present here a systematic study of the feasibility of kilonova detections and assessing the effects of a population of contaminant sources during follow-up of a GW trigger using wide-field optical telescopes. This study will guide the development of observational strategies for wide-field searches that maximize the likelihood of detecting a GW optical counterpart. The paper is organized as follows: In \cref{sec:ch2_contaminants} we describe the kilonova models utilized in this work and develop the list of contaminating transients considered in this work. In \cref{sec:ch2_phase} we  describe observational features that may be used to isolate kilonovae from the population of potential contaminants. In \cref{sec:ch2_MCsims} we present detailed Monte Carlo simulations of observations to investigate the detectability of kilonovae and contaminant rejection for various observational strategies. In \cref{sec:ch2_diff}, we address issues that arise from the possible lack of a deep pre-existing template. We then discuss the effects of modifications to the standard kilonovae models on our results in \cref{sec:ch2_altkilo}. In \cref{sec:ch2_disc}, we asses the expected performance of current and future wide-field telescopes in the framework of kilonova detectability and contaminant rejection and develop strategies for GW follow-up. Conclusions are presented in \cref{sec:ch2_conc}.

@@

\clearpage
\section{Kilonovae and Contaminants}
\label{sec:ch2_contaminants}

@@

@@

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{./figs/chapter2/ch2_f1.pdf}
\caption{Top: Kilonova spectra from \citet{BarnesKasen13} at $\sim$ 1 d and $\sim$ 1 week with ejecta properties of $\Mej =10^{-2}\;M_{\odot}$ and $\beta_{\text{ej}} = 0.2$. Bottom: Light curve for the same model parameters computed at a fiducial distance of 200 Mpc in the DECam {\em r}, {\em i}- and {\em z}-band filters. \vspace{10pt}}
\label{fig:ch2_kilosample}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[trim= 0 0 0 240 ,clip,width=0.75\textwidth]{./figs/chapter2/ch2_f2.pdf}
\caption{Top: {\em i}-band light curves of the contaminants considered in this work at a fiducial distance of 200 Mpc. Bottom: {\em z}-band light curves. See \cref{sec:ch2_WDmerge,sec:ch2_AIC,sec:ch2_ELDD,sec:ch2_type.Ia,sec:ch2_PSFast,sec:ch2_typeIa} for details.}
\label{fig:ch2_LCcont}
\end{figure}

@@
